---
title: "RNA-Seq Data Analysis: Endogenous RAC1 P29S in mouse melanoma GSE118344"
author: "philip.east@crick.ac.uk"
date: 'Compiled: `r format(Sys.time(), "%d %B, %Y @ %H:%M:%S")`'
output:
  html_document:
    df_print: tibble
    toc: true
    toc_depth: 5
    toc_float: true
---

```{css setup_css, echo=FALSE}
body .main-container {
  max-width: 100%;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=15)
```

```{r r_setup,message=FALSE,warning=FALSE}
library( DESeq2 )
library( ggplot2 )
library( limma )
library( xlsx )
options( max.print=300, width=100 )
```

# Introduction

This R markdown document details the analysis of RNA-Seq data
submitted to [GEO
GES118344](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE118344) 
and published in Cancer Cell 36, 1-16 July 8, 2019.

Transgenic mice with a conditional knock-in of the P29S mutation in
the endogenous Rac1 locus were generated and crossed onto C57BL/6J,
Tyr-CreER;BrafCA/wt;Ptenfl/wt mice. Melanomas were induced by topical
4OH-tamoxifen. We compared the gene expression profile in whole tumour
lysates from Tyr-CreER+/-;Ptenfl/wt;BrafCA/wt;Rac1LSL-P29S/wt mice
versus Tyr-CreER+/-;Ptenfl/wt;BrafCA/wt;Rac1wt/wt mice (n = 6 tumours
from 5-6 animals per group).

# Load ExpressionSets

Here we load the data as ExpressionSet objects containing raw
counts and TPMs.

```{r load_data,message=FALSE,warning=FALSE}
load( file = "eset_counts_GSE118344.rda" )
eset_counts
load( file = "eset_TPM_GSE118344.rda" )
eset_TPM
```

# Experimental Design

```{r design,message=FALSE,warning=FALSE}
kable( pData( eset_counts ) )
```

# rlog count data

```{r rlog,message=FALSE,warning=FALSE}
rlog_mat <- DESeq2::rlog( exprs( eset_counts ) )
```

# Filter genes

Remove genes with a zero mean across all samples.

```{r filter,message=FALSE,warning=FALSE}
f <- rowMeans( rlog_mat ) > 0
eset_counts <- eset_counts[ f, ]
eset_TPM <- eset_TPM[ f, ]
rlog_mat <- rlog_mat[ f, ]
```

# DESeq2 Analysis

To determine endogenous RAC1 mutant dependent transcriptional changes
we ran a DESeq2 analysis across RAC1 mutant and wildtype
samples. Differential genes were selected using a 0.05 fdr threshold.

```{r deseq2,message=FALSE,warning=FALSE}
dds <- DESeqDataSetFromMatrix(countData = round( exprs( eset_counts ) ),
                              colData = pData( eset_counts ),
                              design = ~gbg )

dds <- DESeq( dds )
resobj <- results( dds, alpha = 0.05 )
res <- as.data.frame( resobj )
res <- res[ !is.na( res$padj ), ]
decide <- res[ res$padj < 0.05, ]
```

# Replicate group mean

Here we calculate mean log2 normalised counts across replicates.

```{r rgm,message=FALSE,warning=FALSE}
rld <- DESeq2::rlog( dds )
rgm <- do.call( 'cbind', by( t( assay( rld ) ), colData( dds )$gbg, function( mat ) {
    colMeans( mat )
} ) )
rgmTPM <- do.call( 'cbind', by( t( exprs( eset_TPM ) ), eset_TPM$gbg, function( mat ) {
    colMeans( mat )
} ) )
```

# Results

```{r results,message=FALSE,warning=FALSE}
genes <- rownames( decide )
dbase <- data.frame( gene = genes,
                    gene.symbol = fData( eset_counts[ genes, ] )$gene_name,
                    round( rgm[ genes, ], 3 ),
                    TPM = round( rgmTPM[ genes, ], 3 ),
                    log2FoldChange = decide[, "log2FoldChange" ],
                    p.value = decide[ , "pvalue" ],
                    fData( eset_counts[ genes, ] ) )
dbase[ is.na( dbase ) ] <- ""

write.xlsx( dbase, file = "RAC.vs.wt.melanoma.xlsx",
           sheetName = "RAC1vswt", row.names = FALSE )
```

# PCA

```{r pca,message=FALSE,warning=FALSE}
gg <- plotPCA( rld,
              intgroup = "gbg",
              ntop = 1000 ) +
    labs( title = "Sample PCA", color="gbg" )
gg <- gg + geom_label( label = eset_counts$sample.id,
                      alpha = 0.2,
                      nudge_x = 10 )
gg
```

# R Session Info

```{r session,message=FALSE,ywarning=FALSE}
sessionInfo()
```





